---
title: "Project: College Dataset"
author: "Stefano Cattonar, Ines El Gataa, Andrija Nicić, Angelica Rota"
date: "2025-01-90"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r loading libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(corrplot)
library(caret)
```

# Introduction

This project aims to analyze the `College` dataset from *An Introduction to Statistical Learning* to explore the factors that influence the number of applications received by colleges, represented by the response variable `Apps`. The dataset provides comprehensive information on various attributes of U.S. colleges, such as tuition costs, room and board expenses, financial aid, and student demographics. By examining these factors, we seek to better understand the trends and drivers behind college application rates.

# Preprocessing

## Data Examination

```{r}
College = read.csv("https://www.statlearning.com/s/College.csv", header = TRUE)
help(head)
head(College)
```

`X`: name of the college

`Private`: A factor with levels No and Yes indicating private or public university

`Accept`: Number of applications accepted

`Enroll`: Number of new students enrolled

`Top10perc`: Percentage new students from top 10% of high school class

`Top25perc`: Percentage new students from top 25% of high school class

`F.Undergrad`: Number of full time undergraduates

`P.Undergrad`: Number of part time undergraduates

`Outstate`: Out-of-state tuition

`Room.Board`: Room and board costs

`Books`: Estimated book costs

`Personal`: Estimated personal spending

`PhD`: Percentage of faculty with Ph.D.’s

`Terminal`: Percentage of faculty with terminal degree

`S.F.Ratio`: Student/faculty ratio

`perc.alumni`: Percentage alumni who donate

`Expend`: Instructional expenditure per student

`Grad.Rate`: Graduation rate

**Response variable**

`Apps`: Number of applications received


## Data Exploration 

```{r}
summary(College)
```


### Observations

1. We can note that the `X` column will yield no usable information.

```{r}
College <- select(College, -X)
```

2. We have to exclude also `Accept` and `Enroll` to prevent data leakage.

```{r}
College <- select(College, -c(Accept, Enroll))

summary(College)
```

3. All the variables does not have negative values.

4. The variables does not have NA, so we don't need to replace the missing values

5. The variable `Private` is a categorical one so we convert its values.

```{r}
College$Private <- as.factor(College$Private)

College$Private <- as.numeric(College$Private) - 1  # "Yes" is 1, "No" is 0

College$Private[11:30]
```

6. We can note that `Private` is not balanced 



```{r}
table(College$Private)

index_dataset_public <- which(College$Private == 0)

index_dataset_private <- which(College$Private == 1)

private_proportion <- (length(index_dataset_private) / length(index_dataset_public))

private_proportion
```

We can see from private_proportion that "Yes" is 2.65 times more frequent than "No".

# Here we can discuss that Private is unbalanced but 1 to 2 is not a big difference and so we don't need to make corrections to the dataset

7. We want to check if we have some duplicates

```{r}
idx <- which(duplicated(College))
idx
```

## Variable visualization

#### OLD Now we can visualize some of the most interesting variables and their distributions.

#### Now we can visualize the variables and their distributions.

#### Boxplot, excluded Private that are a categorical variable
```{r}

par(mfrow=c(2,3))
for (i in 1:length(College)) {
  if ( names(College[i]) != "Private"){
    boxplot(College[,i], main=names(College[i]))
  }
}
```

###### Boxplot, excluded Private that are a categorical variable

```{r}
par(mfrow=c(2,3))
for (i in 1:length(College)) {
    hist(College[,i], main=names(College[i]))
}
```

OLD We can see that the first two variables are really skewed so we can take the logarithm.

We can see that F.Undergrad and P.Undergrad are really skewed so we can take the logarithm.

```{r}
College$F.Undergrad <- log(College$F.Undergrad)
College$P.Undergrad <- log(College$P.Undergrad)

par(mfrow = c(2, 2))

boxplot(College$F.Undergrad, main="Full time undergraduates")
boxplot(College$P.Undergrad, main= "Part time undergraduates")

hist(College$F.Undergrad, main="Full time undergraduates")
hist(College$P.Undergrad, main= "Part time undergraduates")
```


## Correlation matrix

We plot the correlation matrix to examine the relationships between the predictors and `Apps`

```{r}
correlation_matrix <- cor(College)

corrplot(correlation_matrix, method="color", tl.cex = 0.8)
```

# As you can see from the corralation matrix, there are some variables that are correlated with each other. We can see that the variables Terminal and PhD are really heavily correlated. This fenomena is caused by the fact that PhD containes a sub group of Terminal. So we can exclude one of the two variables, PhD in our case.

# We can also see that the variables Top10perc and Top25perc are correlated. For the same reason we can exclude one of the two variables, Top10perc in our case.

# We can also see that the variables Outstate and Room-and-board are heavy correlated, but it is not a problem because they are not a sub group of each other.




Now we can visualize some of the correlated variables

```{r}
boxplot(College$F.Undergrad ~ College$Private)
boxplot(College$P.Undergrad ~ College$Private)
boxplot(log(College$Apps) ~ College$Private)
```

Now let's normalize the data so that we have all the data with comparable scales.

```{r}
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

college_norm <- College
college_norm[ , 2:ncol(College)] <- lapply(College[ , 2:ncol(College)], normalize)

summary(college_norm)
```

